---
title: "Final Project - Exploratory Data Analysis"
author: "Jessica Foster"
date: "2023-04-23"
output: html_document
---

## Gender

What is the breakdown by gender?
```{r}
almshouse_admissions %>%
  group_by(Gender) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```
767 people admitted to the almshouse were male, 629 female, 4 unknown.

54.8% of all inmates were male. 44.9% were female. 0.3% of unknown gender.

## Description

What is the breakdown by description?
```{r}
almshouse_admissions %>%
  group_by(Description) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```
Most common description is "none indicated" (425), followed by "child" (241), "irish" (141), "black" (95), "deranged" (39), and "pregnant" (36).

30.4% of all inmates with no description, 17.2% described as "child", 10.1% described as "irish", 6.8% described as "black", 2.8% described as "deranged", and 2.6% described as "pregnant".

What is the most common description for males and females, respectively?
```{r}
almshouse_admissions_males <- almshouse_admissions %>%
  filter(Gender == 0) # Just show males

almshouse_admissions_males %>%
  group_by(Description) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```
Most common description for males is "none indicated" (192), followed by "child" (127), "irish" (101), "black" (69), "deranged" (28).

25% of males with no description, 16.6% described as "child", 13.2% described as "irish", 9% described as "black", 3.7% described as "deranged".

```{r}
almshouse_admissions_females <- almshouse_admissions %>%
  filter(Gender == 1) # Just show females

almshouse_admissions_females %>%
  group_by(Description) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```
Most common description for females is "none indicated" (231), followed by "child" (114), "irish" (40), "pregnant" (35), and "black" (28).

36.7% of females with no description, 18.1% described as "child", 6.4% described as "irish", 5.6% described as "pregnant", 4.1% described as "black".

### Summary

To summarize: 

30.4% of all inmates with no description, 17.2% described as "child", 10.1% described as "irish", 6.8% described as "black", 2.8% described as "deranged", and 2.6% described as "pregnant".

25% of males with no description, 16.6% described as "child", 13.2% described as "irish", 9% described as "black", 3.7% described as "deranged".

36.7% of females with no description, 18.1% described as "child", 6.4% described as "irish", 5.6% described as "pregnant", 4.1% described as "black".

So this means a higher percentage of females than males had no description, also higher than percentage of total. Percentage described as "child" about the same for both sexes, with females slightly higher. Greater descrepancy with "irish": higher percentage of males than females, higher also than percentage of total.

Most common descriptions:
1. none indicated
2. child
3. irish
4. black (total), black (male), pregnant (female)
5. deranged (total), deranged (male), black (female)

## Children

How many inmates were children or infants?
```{r}
almshouse_admissions_children <- almshouse_admissions %>%
  filter(grepl("child|foundling|born", Description, ignore.case = TRUE)) 

almshouse_admissions_children %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/1400)
```
22.7% of all inmates were children or infants.

How many children were male and how many were female?
```{r}
almshouse_admissions_children %>%
  group_by(Gender) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count))
```
52.1% of children were male. 47.6% were female.

## Race

```{r}
# almshouse_admissions <- almshouse_admissions %>%
#   mutate(Race = ifelse(Description == "black", "Black"))
```



How many inmates were black?
```{r}
almshouse_admissions_blacks <- almshouse_admissions %>%
  filter(!grepl("blacksmith", Description, ignore.case = TRUE)) %>% # Do not include blacksmiths
  filter(grepl("black|mulatto", Description, ignore.case = TRUE)) 

almshouse_admissions_blacks %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/1400)
```
6.8% described as "black", 9.5% have descriptions containing "black".

```{r}
almshouse_admissions_blacks %>%
  group_by(Description) %>%
  summarize(count = n())
```

## Reason for Discharge (Outcomes)

### All Inmates

What were the most common outcomes for all inmates?
```{r}
almshouse_admissions %>%
  group_by(Reason.for.Discharge) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```
Almost half of all inmates (48.5%) were discharged. 18.7% died. 13.5% "eloped" (ran away). 10.2% were "bound" (indentured). 7.9% with no reason for discharge indicated. Very small number sent to hospital or jail.

### Irish Inmates

What were the most common outcomes for "irish" inmates?
```{r}
almshouse_admissions_irish <- almshouse_admissions %>%
  filter(grepl("irish", Description, ignore.case = TRUE))

almshouse_admissions_irish %>%
  group_by(Reason.for.Discharge) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```
The vast majority (61.7%) of "irish" inmates were discharged, but 19.1% died.

### Black Inmates

What were the most common outcomes for black inmates?
```{r}
almshouse_admissions_blacks %>%
  group_by(Reason.for.Discharge) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```

35.3% of black inmates died. 33.8% were discharged. 10.5% were either "bound" (indentured) or "eloped" (ran away).

```{r}
almshouse_admissions_died <- almshouse_admissions %>%
  filter(Reason.for.Discharge == "died")

almshouse_admissions_died %>%
  group_by(Description) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```

Of the inmates who died, 28.2% had no description, while 12.98% were described as "black", 11.83% as "child", and 10.3% as "irish".

### Children

What were the most common outcomes for children?
```{r}
almshouse_admissions_children %>%
  group_by(Reason.for.Discharge) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))

```
40.1% of children were "bound" (indentured). 32.5% were "discharged". 17.98% died.

### Males

What were the most common outcomes for males?
```{r}
almshouse_admissions_males %>%
  group_by(Reason.for.Discharge) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```
Almost half (48.1%) of male inmates were discharged. 19.4% died. 15.5% "eloped" (ran away). 8.6% were "bound" (indentured).

### Females

What were the most common outcomes for females?
```{r}
almshouse_admissions_females %>%
  group_by(Reason.for.Discharge) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```

Outcomes of female inmates similar to males overall. Almost half (48.97%) of female inmates were discharged. 17.97% died. 12.2% were "bound" (indentured). 10.97% "eloped" (ran away). 9.1% do not have reason indicated.

### Summary

To summarize:

Almost half of all inmates (48.5%) were discharged. 18.7% died. 13.5% "eloped" (ran away). 10.2% were "bound" (indentured). 7.9% with no reason for discharge indicated. Very small number sent to hospital or jail.

Almost half (48.1%) of male inmates were discharged. 19.4% died. 15.5% "eloped" (ran away). 8.6% were "bound" (indentured).

Almost half (48.97%) of female inmates were discharged. 17.97% died. 12.2% were "bound" (indentured). 10.97% "eloped" (ran away). 9.1% do not have reason indicated. Outcomes similar to males overall, though female inmates more likely to be bound out than to run away. Also more likely not to have reason for discharge indicated.

40.1% of children were "bound" (indentured). 32.5% were "discharged". 17.98% died. Children much more likely to be indentured than adult inmates.

The vast majority (61.7%) of "irish" inmates were discharged, but 19.1% died.

35.3% of black inmates died. 33.8% were discharged. 10.5% were either "bound" (indentured) or "eloped" (ran away). Inmates were almost 2x more likely to die if black (35.3% vs. 18.7%).

Of the 262 inmates who died, 12.98% were described as "black".


What were the most common outcomes for inmates with "mental" problem?
```{r}
almshouse_admissions_mental <- almshouse_admissions %>%
  filter(grepl("insane|deranged|lunatic|idiot|foolish|silly", Description, ignore.case = TRUE))

almshouse_admissions_mental %>%
  group_by(Reason.for.Discharge) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```
32.8% of mental inmates "eloped" (ran away). 31.3% were discharged. 15.6% with no reason indicated. 12.5% died. 7.8% sent to hospital.

```{r}
almshouse_admissions_pregnant <- almshouse_admissions %>%
  filter(grepl("pregnant", Description, ignore.case = TRUE))

almshouse_admissions_pregnant %>%
  group_by(Reason.for.Discharge) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```

The vast majority (70.5%) of pregnant inmates were discharged. 13.7% "eloped" (ran away). 11.4% died.

```{r}
almshouse_admissions_old <- almshouse_admissions %>%
  filter(grepl("old", Description, ignore.case = TRUE))

almshouse_admissions_old %>%
  group_by(Reason.for.Discharge) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```
38.7% of "old" inmates were discharged. 32.3% died. 16.1% with no reason indicated. 9.7% "eloped" (ran away).

```{r}
almshouse_admissions_venereal <- almshouse_admissions %>%
  filter(grepl("venereal|veneral", Description, ignore.case = TRUE))

almshouse_admissions_venereal %>%
  group_by(Reason.for.Discharge) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```

```{r}
almshouse_admissions_women_children <- almshouse_admissions %>%
  filter(grepl("child|foundling|born", Description, ignore.case = TRUE) | Gender == 1) 
```

## Bound Out

### Description

Let's look more closely at inmates who were bound out.
```{r}
almshouse_admissions_bound <- almshouse_admissions %>%
  filter(grepl("bound", Reason.for.Discharge, ignore.case = TRUE))

almshouse_admissions_bound %>%
  group_by(Description) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```

Only 10.2% of all inmates were "bound" (indentured) compared to 40.1% of children. This means that an inmate was 4x more likely to be bound out if a child. Also, of all inmates who were bound out, 74% are listed as "child", 9.1% listed as "black child", 3.5% as "born", 2.1% as "mulatto child". This adds up to about 88.7%. So the overwhelming majority of inmates bound out were children.

```{r}
almshouse_admissions_bound %>%
  filter(!grepl("blacksmith", Description, ignore.case = TRUE)) %>% # Do not include blacksmiths
  filter(grepl("black|mulatto", Description, ignore.case = TRUE)) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/143) %>%
  arrange(desc(percentage))
```


### Gender

Let's look at the gender of those bound out.
```{r}
almshouse_admissions_bound %>%
  group_by(Gender) %>%
  summarize(count = n()) %>%
  mutate(percentage =  100*count/sum(count)) %>%
  arrange(desc(percentage))
```
53.84% of inmates bound out were female. 46.2% were male.

### Description & Gender

```{r}
almshouse_admissions_bound %>%
  group_by(Description, Gender) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
```
57 inmates who were bound out were described as "child" and "female". 49 images who were bound out were described as "child" and "male". 8 were "female" "black child" and 5 were "male" "black child".

## Average Length of Stay

```{r}
# Load lubridate library to combine year & month columns
library(lubridate)

# Create function to compare two date variables in dataframe
Date.Check <- function(target.df, date.col1, date.col2) {
  target.df %>% filter(date.col1 <= date.col2)
}

almshouse_admissions_stay <- almshouse_admissions %>%
  filter(YearDischarged != 1) %>%
  mutate(Date.Admitted = make_date(year = Year.Admitted, month = Month.Admitted)) %>% # Combine Year.Admitted and Month.Admitted columns, convert to YYYY-MM-DD format, display in Date.Admitted column
  mutate(Date.Discharged = make_date(year = YearDischarged, month = MonthDischarged)) %>% # Combine YearDischarged and MonthDischarged columns, convert to YYYY-MM-DD format, display in Date.Discharged column
  select(Last.Name, First.Name, Gender, Description, Date.Admitted, Date.Discharged, Reason.for.Discharge) # Select relevant columns
  
# Check dates to make sure Date.Admitted is earlier than or the same as Date.Discharged, filter out rows that don't pass validation
almshouse_admissions_stay <- Date.Check(almshouse_admissions_stay, almshouse_admissions_stay$Date.Admitted, almshouse_admissions_stay$Date.Discharged)

almshouse_admissions_stay <- almshouse_admissions_stay %>% 
   mutate(Length.of.Stay = interval(Date.Admitted, Date.Discharged) %/% months(1)) # Calculate length of stay in almshouse in months

almshouse_admissions_stay %>%
   group_by(Description) %>% # Group by description
   summarize(Average.Length.of.Stay = mean(Length.of.Stay)) %>% # Calculate average length of stay
   arrange(desc(Average.Length.of.Stay))
```
Average length of stay in almshouse for all inmates is 6.02 months. Average length of stay for "child" is 5.2 months. For "black child" it is 7.7 months. Average length of stay for "black" inmates is 4.1 months. Average length of stay for "pregnant" inmates is 6.4 months. Average length of stay for "venereal" inmates is 5.5 months.

```{r}
almshouse_admissions_stay %>%
   group_by(Reason.for.Discharge) %>% # Group by description
   summarize(Average.Length.of.Stay = mean(Length.of.Stay)) %>% # Calculate average length of stay
   arrange(desc(Average.Length.of.Stay))
```

Average length of stay for inmates later bound out is 5.5 months. Average length of stay for inmates discharged is 5.2 months. Average length of stay for inmates who died is 8.6 months.

"irish" inmates make up 9.84% of inmates with year of discharge indicated. They have an average stay of 2.87 months, one of the shortest stays among all inmates.


```{r}
almshouse_admissions_stay_blacks <- almshouse_admissions_blacks %>%
  filter(YearDischarged != 1)

almshouse_admissions %>%
  filter(!grepl("blacksmith", Description, ignore.case = TRUE)) %>% # Do not include blacksmiths
  filter(grepl("black", Description, ignore.case = TRUE) & YearDischarged == 1) %>%
  group_by(Description) %>%
  summarize(count = n()) 
```

12 inmates with "black" in description have no discharge year indicated (133 - 121).

```{r}
almshouse_admissions %>%
  filter(!grepl("blacksmith", Description, ignore.case = TRUE)) %>% # Do not include blacksmiths
  filter(grepl("black|mulatto", Description, ignore.case = TRUE))
```
Nash counts both "black" and "mulatto" as "black", so 159 or 11.4% of total.

```{r}
write.csv(almshouse_admissions, file = "almshouse_admissions.csv", row.names = TRUE, col.names = TRUE)
```

