---
title: "Almshouse Data Preprocessing"
author: "Jessica Foster"
date: "2023-04-23"
output: html_document
---

Load libraries...
```{r}
library(DigitalMethodsData)
library(tidyverse)
library(pdftools)
library(qpdf)
```

First I'm going to load the `almshouse_admissions` dataset from the DigitalMethodsData package.
```{r}
data("almshouse_admissions")
```

Next I'm going to read in the codebook (a Word document) for the Almshouse Admissions data that I downloaded from the original repository: https://repository.upenn.edu/mead/42/. To make the document easier to work with in R, I went ahead and converted the lists of codes to tables in Word. Now we can scrape the content from the Word doc using the `docxtractr` package: https://github.com/hrbrmstr/docxtractr

```{r}
library(docxtractr)

almshouse.codebook <- read_docx("Codebook Almshouse Admissions 1796-1803 - with tables.docx")

# Count the number of tables in the doc
docx_tbl_count(almshouse.codebook)

# Describe each table
docx_describe_tbls(almshouse.codebook)

# Extract the table of description codes and save to variable
description.codes <- docx_extract_tbl(almshouse.codebook, 1, header=TRUE)

# Extract the table of reason for discharge codes and save to variable
reason.for.discharge.codes <- docx_extract_tbl(almshouse.codebook, 2, header=TRUE)
```

Now we have two tables for the codes and can join them to the `almshouse_admissions` dataframe. Let's clean them up a bit first. We'll need to remove the leading zeros from the codes so they match the numbers in the `almshouse_admissions`.
```{r}
description.codes <- description.codes %>%
  mutate(Code = as.numeric(Code))

reason.for.discharge.codes <- reason.for.discharge.codes %>%
  mutate(Code = as.numeric(Code))
```

We can use `left_join()` to join the dataframes, matching up the descriptions and reasons for discharge based on the codes.
```{r}
almshouse_admissions <- left_join(almshouse_admissions, description.codes, by=c("Descriptions.by.Clerk"="Code"))

almshouse_admissions <- left_join(almshouse_admissions, reason.for.discharge.codes, by=c("Reason.Discharged"="Code"))
```

```{r}
# pdf_subset("THE DAILY OCCURRENCE DOCKET OF THE PHILADELPHIA ALMSHOUSE- SELECTED ENTRIES, 1800–1804.pdf", pages = 2:23, output = "daily_occurrence_docket.pdf") # Remove unnecessary pages
# 
# daily_occurrence_docket <- pdf_text("daily_occurrence_docket.pdf") %>%
#    str_split("\n")
# 
# for(i in 1:22) {
#    daily_occurrence_docket[[i]] <- daily_occurrence_docket[[i]][-1:-2]
#    a <- length(daily_occurrence_docket[[i]])
#    b <- a-7
#    daily_occurrence_docket[[i]] <- daily_occurrence_docket[[i]][-b:-a]
# }
#  
# daily_occurrence_docket %>%
#   str_squish() %>%
#   str_split('\", \"\", \"')
```

